(Re-)bridging the anatomy ontologies with SSSOM

### 2024-07-01 presentation.tex
Small fixes to presentation text.

  \note{In response to the limits of the previous system, here are the key
  principles that dictated the design of the new approach.

@@

  \note<1>{Over the past three years, independently of Uberon, a new standard has
  been devised, specifically for the purpose of representing semantic
  mappings: the Simple Standard for Sharing Ontological Mappings.

  The SSSOM standard defines a common data model (formalised in LinkML) to
  represent and manipulate mappings in programming languages. The model
  notably allows to capture a rich set of metadata about each mapping.

  The standard also defines several serialisation formats to store and
  exchange mappings, the most important of which being the ``SSSOM/TSV''
  format, which as the name suggests is based on the ubiquitous TSV format,
  and is intended to make reusing mappings across resources as easy as
  possible.

  You see here an very basic example of what a mapping set in SSSOM/TSV format
  looks like: one mapping per line, each mapping consisting, at its core, of a
  subject, predicate, object triple, associated with several metadata.}

@@

  \note{One of the benefits of the SSSOM standard is that it does not simply
  allow to say that two terms are mapped, to each other.  As we've seen in the
  last example, it also allows to specify an explicit mapping predicate, so as
  to give a precise meaning to each mapping. It is then possible to
  distinguish, for example between mappings of terms that represent exactly
  the same concept in two different vocabularies, and mappings of terms that
  represent \emph{similar}, but not exactly identical, concepts.

  The authors of the SSSOM standard have compiled a dedicated vocabulary, the
  \emph{Semantic Mapping Vocabulary}, to host, among other things, various
  mapping predicates that can be used to describe mappings. In that
  vocabulary, as part of this project, we have added a handful of predicates
  to specifically represent \emph{cross-species mappings}.

@@

  \note{Now that we can store mappings in a dedicated format, we need a way
  to use those mappings when we want to integrate several ontologies together.
  And, as per our third goal, we want to do that in a way that is, as much as
  possible, not specific to Uberon, but reusable by other projects as well.

@@

  The mappings are transformed at build-time into OWL bridge files using a
  custom SSSOM/T-OWL ruleset, that contains rules similar to the one we have
  seen earlier: a rule that says that cross-species mappings between FBbt and
  Uberon are to be transformed into equivalence axioms with a taxonomic
  restriction to the \emph{Drosophila melanogaster} taxon, a rule that says
  that cross-species mappings between XAO and Uberon are to be transformed
  into equivalence axioms with a taxonomic restriction to the \emph{Xenopus}
  taxon, etc.

  That step is performed by a dedicated ROBOT plugins that implement the
  SSSOM/T-OWL language.

  After this step, we have only OWL files, which can all be merged using a
  standard ROBOT merge operation, producing the multi-species ontology that we
  want.}

@@

  \note{For Uberon itself, the main benefit of the new approach is that we now
  have a pipeline that is both more flexible --~for example, easy to modify,
  as all the logic is in the SSSOM/T-OWL ruleset~-- and more robust.

  For downstream users of Uberon, the mappings between Uberon and the
  taxon-specific ontologies are now first-class artefacts. They can have their
  own metadata, allowing for traceability and accountability of each
  individual mapping statement. They can be maintained separately from the
  Uberon ontology itself, for example to delegate them to the taxon-specific
  ontologies themselves (something we have already started doing for the
  mappings with FBbt, which are now maintained by FBbt rather than by Uberon).
  They are available as a distinct release product, which is findable on its
  own, in a format that is still relatively new but that we hope will become
  the standard format for semantic mappings.

>>>>

  \note{In response to the limits of the previous system, we set out to design
  a new system, with the following aims:

@@

  \note<1>{Over the past three years, independently of Uberon, a new standard
  has been devised to work with semantic mappings: the Simple Standard for
  Sharing Ontological Mappings.

  The SSSOM standard defines a common data model to represent and manipulate
  mappings in programming languages. The model notably allows to capture a
  rich set of metadata about each mapping.

  The standard also defines several serialisation formats to store and
  exchange mappings, the most important of which being the ``SSSOM/TSV''
  format, intended to make reusing mappings across resources as easy as
  possible.

  You see here an very basic example of what a mapping set in SSSOM/TSV format
  looks like: one mapping per line, each mapping consisting, at its core, of a
  triple made of a subject, an object, and a mapping predicate. That core is
  associated with several metadata.}

@@

  \note{One of the benefits of the SSSOM standard is that it allows to specify
  an explicit mapping predicate, so as to give a precise meaning to each
  mapping. It is then possible to distinguish, for example between mappings of
  terms that represent exactly the same concept in two different vocabularies,
  and mappings of terms that represent \emph{similar}, but not exactly
  identical, concepts.

  The authors of the SSSOM standard have compiled a dedicated vocabulary, the
  \emph{Semantic Mapping Vocabulary}, to host, among other things, various
  mapping predicates that can be used to qualify mappings. In that vocabulary,
  as part of this project, we have added a handful of predicates to
  specifically represent \emph{cross-species mappings}.

@@

  \note{Now that we can store mappings in a dedicated format, along with their
  metadata and with meaningful mapping predicates, we need a way to use those
  mappings when we want to integrate several ontologies together.  And, as per
  our third goal, we want to do that in a way that is, as much as possible,
  not specific to Uberon, but reusable by other projects as well.

@@

  The bridge files are generated by a ROBOT plugin that implements the
  SSSOM/T-OWL language, and a custom SSSOM/T-OWL ruleset that contains the
  Uberon-specific logic, in the form of rules similar to the one seen in the
  last example.

  After this step, we have only OWL files, which can all be merged using a
  standard ROBOT merge operation, producing the multi-species ontology that we
  want.

  Overall, that pipeline is both more robust, as most of the code is
  consolidated in ROBOT, and more flexible, as the Uberon-specific logic is
  easily modifiable in the SSSOM/T-OWL ruleset.}

@@

  \note{Beyond the more robust and more flexible pipeline, the new approach
  benefits for downstream users of Uberon as well.

  The mappings between Uberon and the taxon-specific ontologies are now
  first-class artefacts. They can have their own metadata, allowing for
  traceability and accountability of each individual mapping statement. They
  can be maintained separately from the Uberon ontology itself, for example to
  delegate them to the taxon-specific ontologies themselves (something we have
  already started doing for the mappings with FBbt, which are now maintained
  by FBbt rather than by Uberon).  They are available as a distinct release
  product, which is findable on its own, in a format that is still relatively
  new but that we hope will become the standard format for semantic mappings.
